---
title: "Visualizing and Modeling 2016 US Presidential Election Results"
output:
  html_document:
    toc: yes
    toc_depth: '4'
  html_notebook:
    theme: yeti
    toc: yes
    toc_depth: 4
    toc_float: yes
---
This data setup stolen brazenly from Jennifer Thompson.
```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

## Load libraries we'll use
# library(devtools)
# install_github('datadotworld/data.world-r')
library(data.world) ## for querying directly from data.world
library(tidyverse)  ## for data wrangling and piping
library(rms)        ## rms has nice functionality for getting predicted values from model objects
library(reshape2)   ## I know this better than tidyr -@rkahne

state_names = c(
'Alabama' = 01L,
'Alaska' = 02L,
'Arizona' = 04L,
'Arkansas' = 05L,
'California' = 06L,
'Colorado' = 08L,
'Connecticut' = 09L,
'Delaware' = 10L,
'Florida' = 12L,
'Georgia' = 13L,
'Hawaii' = 15L,
'Idaho' = 16L,
'Illinois' = 17L,
'Indiana' = 18L,
'Iowa' = 19L,
'Kansas' = 20L,
'Kentucky' = 21L,
'Louisiana' = 22L,
'Maine' = 23L,
'Maryland' = 24L,
'Massachusetts' = 25L,
'Michigan' = 26L,
'Minnesota' = 27L,
'Mississippi' = 28L,
'Missouri' = 29L,
'Montana' = 30L,
'Nebraska' = 31L,
'Nevada' = 32L,
'New Hampshire' = 33L,
'New Jersey' = 34L,
'New Mexico' = 35L,
'New York' = 36L,
'North Carolina' = 37L,
'North Dakota' = 38L,
'Ohio' = 39L,
'Oklahoma' = 40L,
'Oregon' = 41L,
'Pennsylvania' = 42L,
'Rhode Island' = 44L,
'South Carolina' = 45L,
'South Dakota' = 46L,
'Tennessee' = 47L,
'Texas' = 48L,
'Utah' = 49L,
'Vermont' = 50L,
'Virginia' = 51L,
'Washington' = 53L,
'West Virginia' = 54L,
'Wisconsin' = 55L,
'Wyoming' = 56L,
'District of Columbia' = 11L
)

electoral_votes = c(
'Alabama' = 9,
'Alaska' = 3L,
'Arizona' = 11,
'Arkansas' = 6,
'California' = 55L,
'Colorado' = 9,
'Connecticut' = 7,
'Delaware' = 3,
'Florida' = 29,
'Georgia' = 16,
'Hawaii' = 4,
'Idaho' = 4,
'Illinois' = 20,
'Indiana' = 11,
'Iowa' = 6,
'Kansas' = 6,
'Kentucky' = 8,
'Louisiana' = 8,
'Maine' = 4,
'Maryland' = 10,
'Massachusetts' = 11,
'Michigan' = 16,
'Minnesota' = 10,
'Mississippi' = 6,
'Missouri' = 10,
'Montana' = 3,
'Nebraska' = 5,
'Nevada' = 6,
'New Hampshire' = 4,
'New Jersey' = 14,
'New Mexico' = 5,
'New York' = 29,
'North Carolina' = 15,
'North Dakota' = 3,
'Ohio' = 18,
'Oklahoma' = 7,
'Oregon' = 7,
'Pennsylvania' = 20,
'Rhode Island' = 4,
'South Carolina' = 9,
'South Dakota' = 3,
'Tennessee' = 11,
'Texas' = 38,
'Utah' = 6,
'Vermont' = 3,
'Virginia' = 13,
'Washington' = 12,
'West Virginia' = 5,
'Wisconsin' = 10,
'Wyoming' = 3,
'District of Columbia' = 3
)

elections <-list(
  'Election_2000' = read_csv('https://query.data.world/s/404aq0u4n98hmozughp08npg1') %>% select(County, CountyName, StateName, bush, gore, nader, browne, other),
  'Election_2004' = read_csv('https://query.data.world/s/6mr92w4p92rem4bpbs1ggzcui') %>% select(County, CountyName, StateName, bush, kerry, other),
  'Election_2008' = read_csv('https://query.data.world/s/8pz64y5pglv8dt6umysffgahh') %>% select(County, CountyName, StateName, mccain, obama, other),
  'Election_2012' = read_csv('https://query.data.world/s/29sed01wkb9ljmgg0kgj3ztop') %>% select(County, CountyName, StateName, romney, obama, johnson, stein, other),
  'Election_2016' = read_csv('https://query.data.world/s/dmjcapenh4fg8y7chl5oqgjto') %>% select(County, CountyName, StateName, trump, clinton, johnson, stein, other)
)
elections[[1]]$County <- as.integer(elections[[1]]$County)

```

We want to determine what county characteristics may be predictors of both the county's final winner
in the 2016 presidential election, and the margin of victory by which that candidate won.

# Project Setup

## Loading Data

We'll load the county characteristics and 2016 presidential election results datasets directly from data.world.

### County Characteristics

These data are mostly from the 2015 [American Community Survey](https://www.census.gov/programs-surveys/acs/), with additional data from other sources. A
full data dictionary can be found [here](https://github.com/Data4Democracy/election-transparency/blob/master/data-dictionary/county-level/CountyCharacteristics.md).

```{r load_county_data}
library(data.world)

## Set connection (see package README for details: https://github.com/datadotworld/data.world-r)
conn <- data.world('eyJhbGciOiJIUzUxMiJ9.eyJzdWIiOiJwcm9kLXVzZXItY2xpZW50OnJrYWhuZSIsImlzcyI6ImFnZW50OnJrYWhuZTo6ZjE3NDQzZmUtNWM1Yy00ZDllLTgwMGEtMmUwYjhmNTk4ODc2IiwiaWF0IjoxNDg2NTg4MzgzLCJyb2xlIjpbInVzZXJfYXBpX3dyaXRlIiwidXNlcl9hcGlfcmVhZCJdLCJnZW5lcmFsLXB1cnBvc2UiOnRydWV9.txoXoDeTywrXFO5m1-e_8_G7LBOAxjYwds1c9bxyumli0mT9UH16f5PZZjbDaQr60EinxwpuFVU1s7nzJ0uPxw')

# ## What data tables are available? (both dplyr and data.world have a query(); must specify)
# data_list <- data.world::query(conn,
#                                dataset = 'data4democracy/election-transparency',
#                                query = "SELECT * FROM Tables")
# data_list

countyChar <- data.world::query(conn,
                                dataset = 'data4democracy/election-transparency',
                                query = "SELECT * FROM CountyCharacteristics")

```

### Voter Registration

We also want party registration data from November 2016, queried from the full `PartyRegistration`
file. This file includes data pulled from each state's Secretary of State web site. Full data
dictionary is [here](https://github.com/Data4Democracy/election-transparency/blob/master/data-dictionary/county-level/PartyRegistration.md).

Some of the variable names overlap with names in the next dataset; we'll drop variables that are
redundant (state/county names/abbreviations and year/month of registration) and add "Reg" to
everything else except state/county keys to clarify that it's registration info.

```{r load_registration_data}
voterReg2016 <-
  data.world::query(conn,
                    dataset = 'data4democracy/election-transparency',
                    query = "SELECT * FROM PartyRegistration WHERE Year = 2016 AND Month = 11")

voterReg2016 <- voterReg2016 %>%
  select(-one_of("CountyName", "StateName", "StateAbbr", "Year", "Month", "YearMonth"))

names(voterReg2016) <- ifelse(names(voterReg2016) %in% c('State', 'County'), names(voterReg2016),
                              paste0(names(voterReg2016), 'Reg'))

```

### Presidential Election Results by County

These data are collected from a Harvard research project. A full data dictionary can be found [here](https://github.com/Data4Democracy/election-transparency/blob/master/data-dictionary/county-level/PresidentialElectionResults2016.md).

```{r load_results_data}
presResults2016 <- data.world::query(conn,
                                     dataset = 'data4democracy/election-transparency',
                                     query = "SELECT * FROM PresidentialElectionResults2016")

```

## Descriptive Statistics

Let's join the datasets, calculate some proportions, and look at some basic descriptive statistics.

```{r join_data}
## Check what variables are in common
# intersect(names(countyChar), names(voterReg2016))
# intersect(names(countyChar), names(presResults2016))

data2016 <- reduce(list(countyChar, voterReg2016, presResults2016),
                   left_join,
                   by = c('County', 'State'))

## Function to quickly calculate a proportion out of TotalPopulation - we'll need to do this a lot
prop_total <- function(x){ x / data2016$TotalPopulation }

data2016 <- data2016 %>%
  ## Calculate lots of proportion variables
  mutate(propMale = prop_total(Male),
         propKids = prop_total(Age0_4 + Age5_9 + Age10_14 + Age15_19),
         propAdultsNoTeens = 1 - propKids,
         ## 15-19 is included in labor force, marital status questions
         totalAdultsWithTeens = Age15_19 + Age20_24 + Age25_34 + Age35_44 + Age45_54 + Age55_59 +
           Age60_64 + Age65_74 + Age75_84 + Age85,
         propAdultsWithTeens = prop_total(totalAdultsWithTeens),
         ## Only >18 included in education questions
         totalAdultsNoTeens = Age20_24 + Age25_34 + Age35_44 + Age45_54 + Age55_59 + Age60_64 +
           Age65_74 + Age75_84 + Age85,
         propElders = prop_total(Age65_74 + Age75_84 + Age85),
         propNMarried = NeverMarried / totalAdultsWithTeens,
         propHispanic = prop_total(Hispanic),
         propWhite = prop_total(White),
         propBlack = prop_total(Black),
         majWhite = propWhite > 0.5,
         majBlack = propBlack > 0.5,
         propNoHS = (EdK8 + Ed9_12) / totalAdultsNoTeens,
         propHS = EdHS / totalAdultsNoTeens,
         propMoreHS = (EdCollNoDegree + EdAssocDegree + EdBachelorDegree + EdGraduateDegree) /
           totalAdultsNoTeens,
         propMfg2015 = MfgEmp2015 / LaborForce,
         propUnemp = Unemployment / LaborForce,
         propLaborForce = prop_total(LaborForce),
         propStein = stein / totalvotes,
         propJohnson = johnson / totalvotes,
         propTrump = trump / totalvotes,
         propClinton = clinton / totalvotes,
         propVoters = totalvotes / totalAdultsNoTeens,
         votedTrump = rPct > 0.5,
         state_EV = electoral_votes[StateName])
## View full data frame
data2016

```
## Visualizations!
* I made some small modifications to the code above, which mostly turned out to be useless, anyway.
* I (rkahne) really started doing work around here.
* The goal (for me, right now) is to build out dataframes from the great work done by Jennifer Thompson that can be used with ggplot.
* I've done some Very Basic Barplots to get started.  I hope other people can take this stuff, modify it, and find some cool and interesting stuff in here.

###Candidate vote totals by State
```{r}
vote_by_state<-select(data2016, StateName, state_EV, clinton, trump, johnson, stein, other) %>% 
  melt(id= c('StateName','state_EV')) %>% 
  group_by(StateName, state_EV, variable) %>% 
  dplyr::summarize(total_vote = sum(value)) %>% # Dunno how plyr got in, but whatever.
  complete(variable) %>%
  ungroup() %>% 
  group_by(StateName) %>% 
  mutate(sum_total = sum(total_vote, na.rm=T)) %>% 
  mutate(proportion = ifelse(is.na(total_vote),NA,paste0(round(total_vote / sum_total, 4)*100,'%')),
         proportion_numeric = total_vote / sum_total) %>%
  select(-sum_total)

candidate_order <- 'clinton' # CHANGE ME

level_order<-(filter(vote_by_state,variable == candidate_order) %>% 
                     select(StateName, proportion_numeric) %>% 
                     arrange(desc(proportion_numeric)))$StateName 

vote_by_state$StateName<- factor(vote_by_state$StateName, levels = level_order)
vote_by_state$variable<-factor(vote_by_state$variable, levels = c('other','stein','johnson','trump','clinton'))

ggplot(vote_by_state, aes(x=StateName, y=total_vote, fill=variable)) + 
  geom_bar(stat='identity', position = 'fill') +
  scale_fill_manual(values = c('clinton' = 'blue', 'trump' = 'red', 'stein' = 'green','johnson' = 'yellow','other' = 'purple'), 
                    name = 'Candidate',
                    breaks = c('other','stein','johnson','trump','clinton'),
                    labels = c('Others','Jill Stein','Gary Johnson','Donald J Trump','Hillary Clinton')) +
  labs(x = NULL, y = 'Percent of Vote')+
  scale_y_continuous(labels = scales::percent) +
  ggtitle('Percent of Vote by State', subtitle='2016 USA Presidential Election') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
### Candidate Vote Totals by County Within State
Feel free to play with the variable in this chunk.
```{r}
state<-'Tennessee'

vote_by_county<-filter(data2016, StateName == state) %>% 
  select(CountyName, clinton, trump, johnson, stein, other) %>% 
  melt(id= 'CountyName') %>% 
  group_by(CountyName, variable) %>% 
  dplyr::summarize(total_vote = sum(value)) %>% # Dunno how plyr got in, but whatever.
  complete(variable) %>%
  ungroup() %>% 
  group_by(CountyName) %>% 
  mutate(sum_total = sum(total_vote, na.rm=T)) %>% 
  mutate(proportion = ifelse(is.na(total_vote),NA,paste0(round(total_vote / sum_total, 4)*100,'%')),
         proportion_numeric = total_vote / sum_total) %>%
  select(-sum_total)

candidate_order <- 'clinton' # CHANGE ME

level_order<-(filter(vote_by_county,variable == candidate_order) %>% 
                     select(CountyName, proportion_numeric) %>% 
                     arrange(desc(proportion_numeric)))$CountyName 

vote_by_county$CountyName<- factor(vote_by_county$CountyName, levels = level_order)
vote_by_county$variable<-factor(vote_by_county$variable, levels = c('other','stein','johnson','trump','clinton'))

ggplot(vote_by_county, aes(x=CountyName, y=total_vote, fill=variable)) + 
  geom_bar(stat='identity', position = 'fill') +
  scale_fill_manual(values = c('clinton' = 'blue', 'trump' = 'red', 'stein' = 'green','johnson' = 'yellow','other' = 'purple'), 
                    name = 'Candidate',
                    breaks = c('other','stein','johnson','trump','clinton'),
                    labels = c('Others','Jill Stein','Gary Johnson','Donald J Trump','Hillary Clinton')) +
  labs(x = NULL, y = 'Percent of Vote')+
  scale_y_continuous(labels = scales::percent) +
  ggtitle(paste0('Percent of Vote by County - ',state), subtitle='2016 USA Presidential Election') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
###Ternary Plot
* Let's try something crazy?!
* Here's an interesting idea for looking at the county by county results.
    + Utah had a high vote total for Evan McMillin, and this is a way to visualize Utah's uniqueness.
```{r}
library(ggtern)

ternary_vote<-mutate(data2016, third_party = johnson + stein + other) %>% 
  select(County, StateName, clinton, trump, third_party) %>% 
  melt(id= c('County','StateName')) %>% 
  group_by(County, StateName, variable) %>% 
  dplyr::summarize(total_vote = sum(value)) %>% # Dunno how plyr got in, but whatever.
  complete(variable) %>%
  ungroup() %>% 
  group_by(County) %>% 
  mutate(sum_total = sum(total_vote, na.rm=T)) %>% 
  mutate(proportion_numeric = total_vote / sum_total) %>%
  select(-sum_total, -total_vote) %>% 
  spread(key = variable, value = proportion_numeric)

ternary_vote$StateName<-sapply(ternary_vote$StateName, function(i) ifelse(i=='Utah','Utah','Not Utah'))

ggtern(ternary_vote, aes(x=clinton, y=trump, z=third_party, color=StateName))+
  geom_point(aes(alpha = 0.2))

```

###Party Registration by State
Not all states have partisan registration.
```{r}
party_registration <- select(voterReg2016, State, DReg, RReg, OReg, GReg, LReg, NReg) %>% 
  melt(id= 'State') %>% 
  group_by(State, variable) %>% 
  dplyr::summarize(total_reg = sum(value)) %>% 
  complete(variable) %>%
  ungroup() %>% 
  group_by(State) %>% 
  mutate(sum_total = sum(total_reg, na.rm=T)) %>% 
  mutate(proportion = ifelse(is.na(total_reg),NA,paste0(round(total_reg / sum_total, 4)*100,'%')),
         proportion_numeric = total_reg / sum_total) %>%
  select(-sum_total)

party_registration$State<-sapply(party_registration$State, function(i) names(state_names)[which(state_names == i)])

non_partisan_states<-(filter(party_registration, variable == 'NReg') %>% 
                        filter(proportion_numeric == 1))$State
party_registration <- filter(party_registration, !(State %in% non_partisan_states))

party_order <- 'DReg' # CHANGE ME

level_order<-(filter(party_registration,variable == party_order) %>% 
                     select(State, proportion_numeric) %>% 
                     arrange(desc(proportion_numeric)))$State

party_registration$State<- factor(party_registration$State, levels = level_order)
party_registration$variable<-factor(party_registration$variable, levels = c('NReg','OReg','GReg','LReg','RReg','DReg'))

ggplot(party_registration, aes(x=State, y=total_reg, fill=variable)) + 
  geom_bar(stat='identity', position = 'fill') +
  scale_fill_manual(values = c('DReg' = 'blue', 'RReg' = 'red', 'GReg' = 'green','LReg' = 'yellow','OReg' = 'purple','NReg' = 'grey'),
                    name = 'Candidate',
                    breaks = c('NReg','OReg','GReg','LReg','RReg','DReg'),
                    labels = c('No Party','Other Party','Green','Libertarian','Republican','Democratic')) +
  labs(x = NULL, y = 'Percent of Registered Voters')+
  scale_y_continuous(labels = scales::percent) +
  ggtitle('Percent of Voter Registration by State', subtitle='2016') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
Looks like we have a bit of an issue here with Other and Non-partisan registration.  NE and WV are all "Other".

###Partisan Registration by County within State
Feel free to play with the variable in this.
```{r}
state <- 'West Virginia'

party_registration_c <- filter(data2016, StateName == state) %>%  
  select(CountyName, DReg, RReg, OReg, GReg, LReg, NReg) %>% 
  melt(id= 'CountyName') %>% 
  group_by(CountyName, variable) %>% 
  dplyr::summarize(total_reg = sum(value)) %>% 
  complete(variable) %>%
  ungroup() %>% 
  group_by(CountyName) %>% 
  mutate(sum_total = sum(total_reg, na.rm=T)) %>% 
  mutate(proportion = ifelse(is.na(total_reg),NA,paste0(round(total_reg / sum_total, 4)*100,'%')),
         proportion_numeric = total_reg / sum_total) %>%
  select(-sum_total)

party_order <- 'DReg' # CHANGE ME

level_order<-(filter(party_registration_c,variable == party_order) %>% 
                     select(CountyName, proportion_numeric) %>% 
                     arrange(desc(proportion_numeric)))$CountyName

party_registration_c$CountyName<- factor(party_registration_c$CountyName, levels = level_order)
party_registration_c$variable<-factor(party_registration_c$variable, levels = c('NReg','OReg','GReg','LReg','RReg','DReg'))

ggplot(party_registration_c, aes(x=CountyName, y=total_reg, fill=variable)) + 
  geom_bar(stat='identity', position = 'fill') +
  scale_fill_manual(values = c('DReg' = 'blue', 'RReg' = 'red', 'GReg' = 'green','LReg' = 'yellow','OReg' = 'purple','NReg' = 'grey'),
                    name = 'Candidate',
                    breaks = c('NReg','OReg','GReg','LReg','RReg','DReg'),
                    labels = c('No Party','Other Party','Green','Libertarian','Republican','Democratic')) +
  labs(x = NULL, y = 'Percent of Registered Voters')+
  scale_y_continuous(labels = scales::percent) +
  ggtitle('Percent of Voter Registration by County', subtitle=paste0(state,' - 2016')) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

###Voter Turnout by State
* We don't have great data to do this, becuase the census bracket is from age 15-19, which cuts out two years of voting age.
* Maybe because of this (but maybe not!) there are some states where (voted+registered but didn't vote) > total voting population.
    + Even if the issue is not the census populations, it could be merely out of date voter registration data.
```{r}
turnout_data <- group_by(data2016, StateName) %>% 
  dplyr::summarize(total_votes = sum(clinton,trump,johnson,stein,other, na.rm=T),
                   total_reg_no_vote = sum(DReg,RReg,OReg,GReg,LReg,NReg, na.rm=T)-total_votes,
                   total_not_registered = ifelse(sum(TotalPopulation)- sum(Age0_4)- sum(Age5_9)- sum(Age10_14) - sum(Age15_19) -
                     total_reg_no_vote - total_votes<0,NA,sum(TotalPopulation)- sum(Age0_4)- sum(Age5_9)- sum(Age10_14) - 
                     sum(Age15_19) - total_reg_no_vote - total_votes)) %>%  
  melt(id = 'StateName') %>% 
  group_by(StateName, variable) %>% 
  dplyr::summarize(total_pop = sum(value)) %>% 
  complete(variable) %>%
  ungroup() %>% 
  group_by(StateName) %>% 
  mutate(sum_total = sum(total_pop, na.rm=T)) %>% 
  mutate(proportion = ifelse(is.na(total_pop),NA,paste0(round(total_pop / sum_total, 4)*100,'%')),
         proportion_numeric = total_pop / sum_total) %>%
  select(-sum_total)

turnout_data$variable<-factor(turnout_data$variable, levels = c('total_not_registered','total_reg_no_vote','total_votes'))

ggplot(turnout_data, aes(x=StateName, y=total_pop, fill=variable)) + 
  geom_bar(stat='identity', position = 'fill') +
  scale_fill_manual(values = c('total_votes' = 'green', 'total_reg_no_vote' = 'purple', 'total_not_registered' = 'grey'),
                    name = 'Population',
                    breaks = c('total_votes','total_reg_no_vote','total_not_registered'),
                    labels = c('Voted','Registered, but did not vote','Not Registered')) +
  labs(x = NULL, y = 'Percent of Population')+
  scale_y_continuous(labels = scales::percent) +
  ggtitle('Voter Turnout by State', subtitle='2016') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
###Voter Turnout by County Within State
```{r}
state <- 'Kentucky'

turnout_data_c <- filter(data2016, StateName == state) %>% 
  group_by(CountyName) %>% 
  dplyr::summarize(total_votes = sum(clinton,trump,johnson,stein,other, na.rm=T),
                   total_reg_no_vote = sum(DReg,RReg,OReg,GReg,LReg,NReg, na.rm=T)-total_votes,
                   total_not_registered = ifelse(sum(TotalPopulation)- sum(Age0_4)- sum(Age5_9)- sum(Age10_14) - sum(Age15_19) -
                     total_reg_no_vote - total_votes<0,NA,sum(TotalPopulation)- sum(Age0_4)- sum(Age5_9)- sum(Age10_14) - 
                     sum(Age15_19) - total_reg_no_vote - total_votes)) %>%  
  melt(id = 'CountyName') %>% 
  group_by(CountyName, variable) %>% 
  dplyr::summarize(total_pop = sum(value)) %>% 
  complete(variable) %>%
  ungroup() %>% 
  group_by(CountyName) %>% 
  mutate(sum_total = sum(total_pop, na.rm=T)) %>% 
  mutate(proportion = ifelse(is.na(total_pop),NA,paste0(round(total_pop / sum_total, 4)*100,'%')),
         proportion_numeric = total_pop / sum_total) %>%
  select(-sum_total)

turnout_data_c$variable<-factor(turnout_data_c$variable, levels = c('total_not_registered','total_reg_no_vote','total_votes'))

ggplot(turnout_data_c, aes(x=CountyName, y=total_pop, fill=variable)) + 
  geom_bar(stat='identity', position = 'fill') +
  scale_fill_manual(values = c('total_votes' = 'green', 'total_reg_no_vote' = 'purple', 'total_not_registered' = 'grey'),
                    name = 'Population',
                    breaks = c('total_votes','total_reg_no_vote','total_not_registered'),
                    labels = c('Voted','Registered, but did not vote','Not Registered')) +
  labs(x = NULL, y = 'Percent of Population')+
  scale_y_continuous(labels = scales::percent) +
  ggtitle('Voter Turnout by County', subtitle=paste0(state,', 2016')) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


###Victory Margins
* Not quite sure what we are going for, here.

###Demographic / Economic Variables
* Let's look at % of vote faceted over some different variables
* This graphic produces county by county results facetting by quintile of Demographic/Economic variable
    + You can change the variable in the code to see other variables.
* There is one categorical variable, the code for producing it's facetted plot is in commented in this chunk.
```{r}
Demo_Factors <- select(data2016, County, propTrump, NCHS_UrbanRural2013, TotalPopulation, MedianAge, MedianHouseholdIncome, SimpsonDiversityIndex, propWhite, propBlack, propMfg2015, propHS, propMoreHS, propUnemp, propElders, propNMarried)

lm(propTrump~.,select(Demo_Factors,-County)) %>% summary()

Demo_Factors <- select(data2016, County, trump, clinton, johnson, stein, other, NCHS_UrbanRural2013, TotalPopulation, MedianAge, MedianHouseholdIncome, SimpsonDiversityIndex, propWhite, propBlack, propMfg2015, propHS, propMoreHS, propUnemp, propElders, propNMarried)

vote_by_county_f<-select(Demo_Factors, County, clinton, trump, johnson, stein, other) %>% 
  melt(id= c('County')) %>% 
  group_by(County, variable) %>% 
  dplyr::summarize(total_vote = sum(value)) %>% # Dunno how plyr got in, but whatever.
  complete(variable) %>%
  ungroup() %>% 
  group_by(County) %>% 
  mutate(sum_total = sum(total_vote, na.rm=T)) %>% 
  mutate(proportion = ifelse(is.na(total_vote),NA,paste0(round(total_vote / sum_total, 4)*100,'%')),
         proportion_numeric = total_vote / sum_total) %>%
  select(-sum_total)

vote_by_county_f <- left_join(vote_by_county_f, Demo_Factors, by='County')

candidate_order <- 'clinton' # CHANGE ME

level_order<-(filter(vote_by_county_f,variable == candidate_order) %>% 
                     select(County, proportion_numeric) %>% 
                     arrange(desc(proportion_numeric)))$County 

vote_by_county_f$County<- factor(vote_by_county_f$County, levels = level_order)
vote_by_county_f$variable<-factor(vote_by_county_f$variable, levels = c('other','stein','johnson','trump','clinton'))

# ggplot(filter(vote_by_county_f, !is.na(NCHS_UrbanRural2013)), aes(x=County, y=total_vote, fill=variable)) + 
#   geom_bar(stat='identity', position = 'fill') +
#   scale_fill_manual(values = c('clinton' = 'blue', 'trump' = 'red', 'stein' = 'green','johnson' = 'yellow','other' = 'purple'), 
#                     name = 'Candidate',
#                     breaks = c('other','stein','johnson','trump','clinton'),
#                     labels = c('Others','Jill Stein','Gary Johnson','Donald J Trump','Hillary Clinton')) +
#   labs(x = NULL, y = 'Percent of Vote')+
#   scale_y_continuous(labels = scales::percent) +
#   ggtitle('Percent of Vote by County', subtitle='2016 USA Presidential Election') +
#   theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
#   facet_wrap(~NCHS_UrbanRural2013, scales = 'free_x')

facetting_variable <- 'SimpsonDiversityIndex' #Change Me

vote_by_county_f$fct <- cut(vote_by_county_f[[facetting_variable]], 
                            breaks = c(0,quantile(vote_by_county_f[[facetting_variable]], 
                                              probs = c(0.1, 0.25, 0.5, 0.75, 0.9),
                                              na.rm = T),Inf),
                            ordered_result = T,
                            labels = c('Lowest 10%','10th - 25th %ile',
                                       '25th %ile - Median','Median - 75th %ile',
                                       '75th - 90th %ile', 'Top 10%'))

ggplot(filter(vote_by_county_f, !is.na(fct)), aes(x=County, y=total_vote, fill=variable)) + 
  geom_bar(stat='identity', position = 'fill') +
  scale_fill_manual(values = c('clinton' = 'blue', 'trump' = 'red', 'stein' = 'green','johnson' = 'yellow','other' = 'purple'), 
                    name = 'Candidate',
                    breaks = c('other','stein','johnson','trump','clinton'),
                    labels = c('Others','Jill Stein','Gary Johnson','Donald J Trump','Hillary Clinton')) +
  labs(x = NULL, y = 'Percent of Vote')+
  scale_y_continuous(labels = scales::percent) +
  ggtitle(paste0('Percent of Vote by County ~ ',facetting_variable), subtitle='2016 USA Presidential Election') +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  facet_wrap(~fct, scales = 'free_x')


```

