---
title: "Add_Demos"
author: "Chris Dick"
date: "January 28, 2017"
output: 
  md_document:
    variant: markdown_github
---
# Add Demographic Variables to Registration and Voting Data

The code below is an example of adding demographic data to our voting and registration data. At this point I have only added age by race variables from the 2015 5-year American Community Survey data. However, the code below can be relatively easily adapted to pull anything else from the ACS.

## What you will need to get started:
- A Census API Token, if you don't already have one it is super easy. Go here to request: http://api.census.gov/data/key_signup.html
- To use the below code you will need the updated uselections R package, which can be downloaded here using the devtools package and the command `devtools::install_github("Data4Democracy/election-transparency/r-packages/uselections")`

```{r, results='hide', message=FALSE}
library(dplyr, quietly = TRUE)
library(uselections, quietly = TRUE)  # I am pulling the data from the R package, could use data.world
library(knitr, quietly = TRUE)
library(acs, quietly = TRUE)
library(reshape2, quietly = TRUE)
library(stringr, quietly = TRUE)
library(tidyr, quietly = TRUE)
```

## Pulling and cleaning the data

First we need to pull the data from the Census API. The below code pulls data at the County level by age, race, sex, and Hispanic origin.

```{r acs, results='hide', message=FALSE}

api.key.install(key = '378bfe7e9514b505b44c7aff5702dc081046b38b') # You will need to enter an API key

geokey <- geo.make(state = "*", 
                   county = "*")

# Set list of attributes you want to pull, you will want to look for the table numbers. 


pull_list <- c("B01001A",    # Sex by Age: White Alone
               "B01001B",    # Sex by Age: Black Alone
               "B01001C",    # Sex by Age: 
               "B01001D",    # Sex by Age:
               "B01001E",
               "B01001F",
               "B01001G",
               "B01001H",    # Sex by Age: White Alone Non-Hispanic
               "B01001I")

# Define function to pull data and dump into a data frame
pull <- function(pull){
  df <- acs.fetch(endyear = 2015, span = 5, geography = geokey, table.number = pull, 
            col.name = "pretty")
  new_df <- data.frame(paste0(str_pad(df@geography$state, 2, "left", pad="0"), 
                               str_pad(df@geography$county, 3, "left", pad="0")), 
                        df@estimate, 
                        stringsAsFactors = FALSE)
  names(new_df)[1] <- 'geoid'
  return(new_df)
}

#Apply this function over our list of tables, the result will be a list of DFs
dfs <- lapply(pull_list, pull)

# split out the non-hispanic white alone because it does not have the same number of categories
# from the API
df_nhwa <- dfs[[8]]
dfs <- dfs[-8]

demo_frame <- data.frame(NULL)

# Join together all of your tables. I am doing it this way because all of my tables have the same 
# dimensions and are measured the same way. If I had tables with different demographic dimensions
# I would want to clean them separately, so group and join your tables accordingly.
for (i in 1:length(dfs)){
  if (i == 1){
    demo_frame <- dfs[[i]]
  } else{
    demo_frame <- demo_frame %>% inner_join(dfs[[i]], by = "geoid")
  }
}
```

Now that we have all of our data in on big data frame with a ton of columns, I want to work on getting something more usable. For these age, sex, race, Hispanic origin data I will give us the component parts to be able to calculate other important variables. Accordingly, I leave the data separated by race / Hispanic origin, but group age into 2 categories:

1. Under 18 Years Old
2. Over 18 Years Old (Voting Age Population)

We may want to make different decisions in the future, but I think this is a good start.

```{r clean, results='hide', message=FALSE}
# Start with the data set with everything but the Non-Hispanic White Alone data
demo_frame1 <- demo_frame %>%
  melt(id = "geoid") %>%
  separate(variable, into = c("variable", "race", "sex", "age"), fill = "right", sep = "\\.\\.") %>%
  select(-2) %>%
  filter(!is.na(age)) %>%
  dcast(geoid + race + sex ~ age, value.var = "value") %>%
  mutate(under18 = Under.5.years + `5.to.9.years` + `10.to.14.years` + `15.to.17.years`,
         votingAge = `18.and.19.years` + `20.to.24.years` + `25.to.29.years` +
           `30.to.34.years` + `35.to.44.years` + `45.to.54.years` + `55.to.64.years` 
         + `65.to.74.years` + `75.to.84.years` + `85.years.and.over`) %>%
  select(-(4:17)) %>%
  mutate(sex = ifelse(sex == ".Female", "female", "male"),
         race = ifelse(race == "AMERICAN.INDIAN.AND.ALASKA.NATIVE.ALONE", "aian",
                ifelse(race == "ASIAN.ALONE", "asian", 
                ifelse(race == "BLACK.OR.AFRICAN.AMERICAN.ALONE", "black",
                ifelse(race == "HISPANIC.OR.LATINO", "hisp",
                ifelse(race == "NATIVE.HAWAIIAN.AND.OTHER.PACIFIC.ISLANDER.ALONE", "nhpi",
                ifelse(race == "SOME.OTHER.RACE.ALONE", "sor",
                ifelse(race == "TWO.OR.MORE.RACES", "multi", "white")))))))) %>%
  melt() %>%
  dcast(geoid + race + variable ~ sex) %>%
  mutate(pop = male + female) %>%
  select(-(4:5)) %>%
  dcast(geoid ~ race + variable)

# Clean the Non-Hispanic White Alone data
demo_frame2 <- df_nhwa %>%
  melt(id = "geoid") %>%
  separate(variable, into = c("variable", "race", "race2", "sex", "age"), 
           fill = "right", sep = "\\.\\.") %>%
  select(-2, -4) %>%
  filter(!is.na(age)) %>%
  dcast(geoid + race + sex ~ age, value.var = "value") %>%
  mutate(under18 = Under.5.years + `5.to.9.years` + `10.to.14.years` + `15.to.17.years`,
         votingAge = `18.and.19.years` + `20.to.24.years` + `25.to.29.years` +
           `30.to.34.years` + `35.to.44.years` + `45.to.54.years` + `55.to.64.years` 
         + `65.to.74.years` + `75.to.84.years` + `85.years.and.over`) %>%
  select(-(4:17)) %>%
  mutate(sex = ifelse(sex == ".Female", "female", "male"),
         race = "nhwa") %>%
  melt() %>%
  dcast(geoid + race + variable ~ sex) %>%
  mutate(pop = male + female) %>%
  select(-(4:5)) %>%
  dcast(geoid ~ race + variable) %>%
  inner_join(demo_frame1, by = "geoid") %>%
  filter(as.numeric(geoid) < 72000)
```

The race categories are labelled as:

1. nhwa: Non-Hispanic White Alone
2. aian: American Indian or Alaska Native
3. asian: Asian
4. black: Black or African American
5. hisp: Hispanic or Latino (Any Race)
6. multi: 2 or More Races
7. nhpi: Native Hawaiian or other Pacific Islander
8. sor: Some other Race
9. white: White

To get the total population sum the following categories:

* aian
* asian
* black
* multi
* nhpi
* sor
* white

