---
title: "ReadCongressionalData1980to2014"
author: "Chris Dick"
date: "February 7, 2017"
output: html_document
---

Script used to read congressional election results for 1980 to 2004 from the [Constituency-Level Elections Archive](http://www.electiondataarchive.org/datacenter.html)

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(reshape2)

# Load state info from data.world (for state Postal and FIPS codes) and results downloaded from CLEA
# Citation for data: Kollman, Ken, Allen Hicken, Daniele Caramani, David Backer, and David Lublin. 
# 2016. Constituency-Level Elections Archive. Produced and distributed by Ann Arbor, MI: 
# Center for Political Studies, University of Michigan.
# Downloaded on Feb-06-2017 from: http://www.electiondataarchive.org/datacenter.html
df_state <- read.csv("https://query.data.world/s/7g2b95zv0avmvfw6gqyucbxdp",header=T)
load("~/Desktop/clea_20161024.Rdata")
```

Clean up the data for our use. One major issue we have is that any race that was uncontested is simply coded as -992. Therefore, for states that had uncontested elections in any of the years, you **cannot** sum the votes to the state level and get the correct state total of votes. *This is something we need to fix in the future*.

```{r}

regexp <- "[[:digit:]]+"
us_df <- clea_20161024 %>%
  filter(ctr_n == "US" & yr > 1979) %>%
  select((5:9), 11:12, 16, 19, 20, 32) %>%
  mutate(dist = str_extract(cst_n, regexp),
         sub = tolower(sub)) %>%
  select(-(4:5))

df_state2 <- df_state %>%
  mutate(State = ifelse(State < 10, paste0("0", State), State),
         StateName = tolower(StateName)) %>%
  select(1:3)

df_final <- us_df %>%
  left_join(df_state2, by = c("sub" = "StateName")) %>%
  mutate(party = ifelse(pty == 180, "D", 
                 ifelse(pty == 583, "R",
                 ifelse(pty == 409, "L",
                 ifelse(pty == 252, "G", "O")))),
         State = as.character(State)) %>%
  select(1, 2, 11, 12, 10, 13, 6, 7) %>%
  rename(year = yr,
         month = mn,
         state_fips = State,
         state = StateAbbr,
         district = dist,
         total_votes = vv1,
         votes = cv1) %>%
  mutate(district = ifelse(is.na(district), 0, as.numeric(district))) %>%
  dcast(year + month + state_fips + state + district + total_votes ~ party, value.var = "votes", 
        fun.aggregate = sum) %>%
  mutate(dPct = D / total_votes,
         rPct = R / total_votes,
         oPct = (G + L + O) / total_votes) %>%  # This includes all non-D/R parties, not just "other"
  arrange(year, state_fips, district)

# Filter uncontested elections because data is not reported for these.
# We need to get data for these races from another source to have correct state totals.
df_unc <- df_final %>% filter(total_votes == -992)

fwrite(df_final, "./HouseElectionResults1980to2014.csv")

```